Supplementary Information

### 2021-03-18 doc/manuscript/ForestGEO_treerings_SI.Rmd
remove unnecessary latex
#121

climwin_generic = "Rows correspond to the four different models run: (1) all $RW$ data, (2) $RW$ for the subset of cores for which $DBH could be reconstructed, (3) $BAI$, and (4)  AGB growth. The first four columns give statistics for time windows tested in *climwin*, where window open and close indicate months prior to the last month in the climwin analysis window for the site (Table S1). Difference in $AIC$ relative to a model with no climate term (first column) is used to select the optimal time window (second column). The third and fourth columns give values of linear and quadratic terms for each time window. The final column shows the correlation of individual-level residuals to the selected climate variable, with the function fit in *climwin* (not the final fit)."

@@

$$DBH_Y  = DBH_{m} - 2*\left[r_{bark,m} - r_{bark,Y} + \sum_{y=Y}^{m} RW_{y} \right]$$
Here, $r_{bark}$ denotes bark thickness, which was estimated on a species-specific basis from allometries (Tables `r species_table`, `r bark_allometries`).
If $DBH_m$ was taken before coring, $RW$ values between $y=m$ and $y=Y$ were added instead of subtracted.
If $DBH_m$ was taken $\le$ 10 years after the last year of measurement in the core, we used the equation above, assuming that the $n$ missing measurements between the year of the last $RW$ record in the core and year $m$ were equal to the last $n$ records in the core.

In some cases, negative values of $DBH_Y$ were predicted for the earliest years in the core record; that is, the sum of $RW$ from the earliest year in the core record to year $m$, plus the predicted change in $r_{bark}$, was greater than half of $DBH_m$. 
Such cases could occur due to a measurement or data error, or if the tree's growth was asymmetrical around the pith, as commonly occurs [@speer_fundamentals_2010]. 
In cases where the error was egregious -- i.e., where predicted $DBH_Y$ of the earliest year in the record divided by $DBH_m$ was < -0.5 -- we removed the tree from the analysis.
We assumed that smaller errors were attributable to asymmetrical growth, and re-distributed the error evenly across years so that $DBH_Y$ prior to the first year in the core was zero. 
This correction was applied only to $DBH_Y$, and not to any $RW$ records.

@@

Prior to analysis, data were prepared and cleaned as described in the Methods section, resulting in an identical set of records for input into each analysis.
For the approach developed here, analysis was conducted as described in the Methods section, but with the *climwin* climate variable selection process limited to just the species of interest (as opposed to all species at the site), climate variables considered individually rather than additively, analysis of only first-order linear relationships, and with start date adjusted to match the conventional method (see below).
*Climwin* is a useful analytical tool that automates the process of running and comparing numerous regressions, but its use does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and time windows can be compared without climwin (e.g., REFS), and *climwin* can also be used on residual chronologies. 
Following the *climwin* analysis step, we extracted *beta* coefficients describing the slope of the relationship between climate and $RW$.
*Beta* coefficients, along with their standard error, were obtained for each month within the analysis time frame (Table `r site_details`) and for the time window identified as optimal by *climwin*.

@@

*Beta* (slope) coefficients for the relationship between tree growth and the monthly climate variable were derived as in @helcoski_growing_2019.
Briefly, we used the R packages *dplR* [@bunn_dendrochronology_2008] and *bootRes* [@zang_dendroclimatic_2013] to analyze relationships between $RW$ and monthly climate following the bootstrapping methodology of [@biondi_dendroclim2002_2004].
Pearson correlations between climate variables and tree-ring chronologies were converted to linear slopes using the method of @charney_observed_2016.

@@

Our analysis included two sites where climate change has had pronounced effects on tree growth: Little Tesuque, New Mexico, USA (LT) and Scotty Creek, NW Territories, Canada (SC).
At LT, increasingly warm drought has dramatically reduced growth [@touchan_millennial_2011; @williams_temperature_2013], resulting in many missing rings in recent years.
At SC, rapidly rising temperatures are causing melting permafrost, summer moisture stress, resulting in negative growth trends in basal area index ($BAI$) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].

@@

The second approach involved detrending the climate variables using a GAM with automated selection of the number of knots prior to the *climwin* step.
Thus, this approach correlates growth anomalies to climate anomalies around a more slowly changing mean, and using it to select climate driver variables inherently assumes that the most important climate drivers of high-frequency (annual) growth variation are also the most important drivers of longer-term growth changes in response to changing climate.
For PPT at LT, this approach and our baseline method identified identical time windows for all growth metrics. 
For the temperature variable at LT, this approach and our baseline method consistently identified $T_{max}$ as the most important variable, but selected slightly different time windows with $RW$ the response metric (May-July for baseline method, April-July when $T_{max}$ was detrended). 
For the water variable at SC, the two methods identified identical climate drivers (previous May or June - previous August PPT, depending on growth metric but not the analysis method).
For the temperature variable at SC, the methods identified different variables and windows, with the baseline method selecting July $T_{max}$ and the detrending method selecting previous December $T_{mean}$ (which makes little sense biologically). 
For the water variable at SCBI, this approach and our baseline method consistently identified previous June- current August PPT as the strongest climate driver, with the exception that current May-August came out as the strongest driver in the detrended analysis for $\Delta AGB$. 
For the temperature variable at SCBI, results were somewhat different, with this approach identifying $T_{max}$ as the most important variable, and the baseline approach identifying PET. Both identified time windows within the current growing season when $RW$ was the metric, but time windows were not similar when $BAI$ or AGB was the metric.
Given little meaningful difference between these approaches, and no clear advantage of detrending the climate variable, we preferred the baseline method.

The third approach involved splitting analyses into decades before and after 1970 [*sensu*  @sniderhan_growth_2016] -- an approach that has also been applied in other studies in rapidly warming climates [e.g., @lloyd_no_2002; @driscoll_divergent_2005;@wilmking_recent_2004].
This approach should accurately capture the most important climatic drivers of growth variation prior to the acceleration of climate change in recent decades, and should also identify the drivers of recent high-frequency growth variation.
For PPT at LT, optimal time windows pre-1970 and post-1970 were similar to those for the entire time period.
For the temperature variable at LT, $T_{max}$ was consistently identified as the most important variable, but it's optimal time window extended back to the previous growing season pre-1970, but was limited to the current growing season post-1970 and for the analysis as a whole.
For the water variable at SC, PPT was consistently selected as a better predictor than PDF, and responses pre- and post-1970 were generally similar to those over the full time period.
For the temperature variable at SC, the pre- and post- 1970 time periods both identified $T_{min}$ temperatures during a single month in the previous summer as the optimal climate driver for $RW$, whereas the analysis for the full time period identified $T_{max}$ of current July (Fig. `r SC_1970break`).
For the water variable at SCBI, PPT was consistently selected as a better predictor than PDF, with similar time windows pre-1970 and for the full analysis period, compared to only a weak effect over a different time period (current May) post-1970 (Fig. `r SC_1970break`).
For the temperature variable at SCBI, selected variables (PET, $T_{max}$, or $T_{mean}$) and time windows were variable across the three analysis periods, with only weak relationships in the post-1970 analysis period.
These results indicate potentially shifting climate sensitivities, particularly at SC (Fig. `r SC_1970break`), which may not be adequately captured by our baseline method.
Further analysis into how climate sensitivities may be shifting would be valuable, but is beyond the scope of this analysis.

>>>>

climwin_generic = "Rows correspond to the four different models run: (1) all RW data, (2) RW for the subset of cores for which $DBH could be reconstructed, (3) BAI, and (4)  AGB growth. The first four columns give statistics for time windows tested in *climwin*, where window open and close indicate months prior to the last month in the climwin analysis window for the site (Table S1). Difference in $AIC$ relative to a model with no climate term (first column) is used to select the optimal time window (second column). The third and fourth columns give values of linear and quadratic terms for each time window. The final column shows the correlation of individual-level residuals to the selected climate variable, with the function fit in *climwin* (not the final fit)."

@@

$$DBH_Y  = DBH_{m} - 2*\left[r_{bark,m} - r_{bark,Y} + \sum_{y=Y}^{m} RW_{y} \right]$$
Here, $r_{bark}$ denotes bark thickness, which was estimated on a species-specific basis from allometries (Tables `r species_table`, `r bark_allometries`).
If $DBH_m$ was taken before coring, RW values between $y=m$ and $y=Y$ were added instead of subtracted.
If $DBH_m$ was taken $\le$ 10 years after the last year of measurement in the core, we used the equation above, assuming that the $n$ missing measurements between the year of the last RW record in the core and year $m$ were equal to the last $n$ records in the core.

In some cases, negative values of $DBH_Y$ were predicted for the earliest years in the core record; that is, the sum of RW from the earliest year in the core record to year $m$, plus the predicted change in $r_{bark}$, was greater than half of $DBH_m$. 
Such cases could occur due to a measurement or data error, or if the tree's growth was asymmetrical around the pith, as commonly occurs [@speer_fundamentals_2010]. 
In cases where the error was egregious -- i.e., where predicted $DBH_Y$ of the earliest year in the record divided by $DBH_m$ was < -0.5 -- we removed the tree from the analysis.
We assumed that smaller errors were attributable to asymmetrical growth, and re-distributed the error evenly across years so that $DBH_Y$ prior to the first year in the core was zero. 
This correction was applied only to $DBH_Y$, and not to any RW records.

@@

Prior to analysis, data were prepared and cleaned as described in the Methods section, resulting in an identical set of records for input into each analysis.
For the approach developed here, analysis was conducted as described in the Methods section, but with the *climwin* climate variable selection process limited to just the species of interest (as opposed to all species at the site), climate variables considered individually rather than additively, analysis of only first-order linear relationships, and with start date adjusted to match the conventional method (see below).
*Climwin* is a useful analytical tool that automates the process of running and comparing numerous regressions, but its use does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and time windows can be compared without climwin (e.g., REFS), and *climwin* can also be used on residual chronologies. 
Following the *climwin* analysis step, we extracted *beta* coefficients describing the slope of the relationship between climate and RW.
*Beta* coefficients, along with their standard error, were obtained for each month within the analysis time frame (Table `r site_details`) and for the time window identified as optimal by *climwin*.

@@

*Beta* (slope) coefficients for the relationship between tree growth and the monthly climate variable were derived as in @helcoski_growing_2019.
Briefly, we used the R packages *dplR* [@bunn_dendrochronology_2008] and *bootRes* [@zang_dendroclimatic_2013] to analyze relationships between RW and monthly climate following the bootstrapping methodology of [@biondi_dendroclim2002_2004].
Pearson correlations between climate variables and tree-ring chronologies were converted to linear slopes using the method of @charney_observed_2016.

@@

Our analysis included two sites where climate change has had pronounced effects on tree growth: Little Tesuque, New Mexico, USA (LT) and Scotty Creek, NW Territories, Canada (SC).
At LT, increasingly warm drought has dramatically reduced growth [@touchan_millennial_2011; @williams_temperature_2013], resulting in many missing rings in recent years.
At SC, rapidly rising temperatures are causing melting permafrost, summer moisture stress, resulting in negative growth trends in basal area index (BAI) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].

@@

The second approach involved detrending the climate variables using a GAM with automated selection of the number of knots prior to the *climwin* step.
Thus, this approach correlates growth anomalies to climate anomalies around a more slowly changing mean, and using it to select climate driver variables inherently assumes that the most important climate drivers of high-frequency (annual) growth variation are also the most important drivers of longer-term growth changes in response to changing climate.
For PPT at LT, this approach and our baseline method identified identical time windows for all growth metrics. 
For the temperature variable at LT, this approach and our baseline method consistently identified $T_{max}$ as the most important variable, but selected slightly different time windows with RW the response metric (May-July for baseline method, April-July when $T_{max}$ was detrended). 
For the water variable at SC, the two methods identified identical climate drivers (previous May or June - previous August PPT, depending on growth metric but not the analysis method).
For the temperature variable at SC, the methods identified different variables and windows, with the baseline method selecting July $T_{max}$ and the detrending method selecting previous December $T_{mean}$ (which makes little sense biologically). 
For the water variable at SCBI, this approach and our baseline method consistently identified previous June- current August PPT as the strongest climate driver, with the exception that current May-August came out as the strongest driver in the detrended analysis for $\Delta$AGB. 
For the temperature variable at SCBI, results were somewhat different, with this approach identifying $T_{max}$ as the most important variable, and the baseline approach identifying PET. Both identified time windows within the current growing season when RW was the metric, but time windows were not similar when BAI or $\Delta$AGB was the metric.
Given little meaningful difference between these approaches, and no clear advantage of detrending the climate variable, we preferred the baseline method.

The third approach involved splitting analyses into decades before and after 1970 [*sensu*  @sniderhan_growth_2016] -- an approach that has also been applied in other studies in rapidly warming climates [e.g., @lloyd_no_2002; @driscoll_divergent_2005;@wilmking_recent_2004].
This approach should accurately capture the most important climatic drivers of growth variation prior to the acceleration of climate change in recent decades, and should also identify the drivers of recent high-frequency growth variation.
For PPT at LT, optimal time windows pre-1970 and post-1970 were similar to those for the entire time period.
For the temperature variable at LT, $T_{max}$ was consistently identified as the most important variable, but it's optimal time window extended back to the previous growing season pre-1970, but was limited to the current growing season post-1970 and for the analysis as a whole.
For the water variable at SC, PPT was consistently selected as a better predictor than PDF, and responses pre- and post-1970 were generally similar to those over the full time period.
For the temperature variable at SC, the pre- and post- 1970 time periods both identified $T_{min}$ temperatures during a single month in the previous summer as the optimal climate driver for RW, whereas the analysis for the full time period identified $T_{max}$ of current July (Fig. `r SC_1970break`).
For the water variable at SCBI, PPT was consistently selected as a better predictor than PDF, with similar time windows pre-1970 and for the full analysis period, compared to only a weak effect over a different time period (current May) post-1970 (Fig. `r SC_1970break`).
For the temperature variable at SCBI, selected variables (PET, $T_{max}$, or $T_{mean}$) and time windows were variable across the three analysis periods, with only weak relationships in the post-1970 analysis period.
These results indicate potentially shifting climate sensitivities, particularly at SC (Fig. `r SC_1970break`), which may not be adequately captured by our baseline method.
Further analysis into how climate sensitivities may be shifting would be valuable, but is beyond the scope of this analysis.
